---
title: "Post of Emphasis on: Points Of Emphasis"
author: "robert"
date: 2024-03-15
category: [referees]
---

```{r start, include=FALSE}
library(tidyverse)

poe_data <- read_csv("poe_2023_24.csv")
```

I have to admit I do not know how to adequately explain the portion of the NBA's online website that this post is focused on. It is the [NBA Official site](https://official.nba.com/) which focuses on the Officials of the NBA -- also known as referees. This is not to be confused with the larger [NBA dot com](https://www.nba.com) which is described as:

> "The **official** site of the National Basketball Association. Follow the action on NBA scores, schedules, stats, news, Team and Player news."

in web searches. That is the official NBA site, of which the NBA Official site is within that, which I guess makes it the official NBA Official site?

ANYWAY: the TLDR is Tom and Ethan are both probably right and there's always been murkiness with NBA refereeing that I give you a bunch of background on as well as concrete data about.

# Motivation

A little bit of expose is that Tom Haberstroh wrote a fascinating article, [NBA scoring is crashing. Why?](https://www.tomthefinder.com/p/nba-scoring-is-crashing-why), that noticed a sharp dip in NBA scoring rates that started around late January of this Season. Tom also pointed out that the free throw rate in February declined by about 10% which would likely be the largest factor for why scoring started to decline. Ethan Strauss then piggy-backed off of Tom's article with some reporting about NBA league circles on his Substack, [General Managers and Coaches Think the NBA Changed Its Rules Midseason](https://www.houseofstrauss.com/p/general-managers-and-coaches-think), to flesh out the why for the NBA's recent decline in scoring. The gist of Ethan's article is that NBA team personnel believe a change in how NBA Official's referee a game has changed *and that the teams were not notified of this change*. Tom subsequently followed up on his initial article, [The NBA's bizarre free throw drought drought continues: Is it the NBA's pitch clock?](https://www.tomthefinder.com/p/the-nbas-bizarre-free-throw-drought), by digging deeper into officiating related changes that occurred in February. Along with a decline in the fouls called by referees declining, Tom also noticed that the average game-length was shortening and that betting game totals began heavily hitting the "unders."

Undoubtedly the results Tom found are true. And Ethan's reporting about how NBA personnel feel towards changes in refereeing is not really falsifiable, but more importantly no real reason to not believe him. But one aspect that is not so clear is if teams were not notified of this change. Which brings me to some officiating minutia. Details, if you will, about how NBA officiating operates, has operated, and some statistics about their examples in describing how games should be officiated.

# Points Of Emphasis

The NBA's Official account will give guidance to the public at large on [Points of Emphasis](https://official.nba.com/2023-24-points-of-emphasis/) (POE). This topic was mildly brought up in Ethan Strauss's [podcast with Tom Haberstroh](https://www.houseofstrauss.com/p/hos-tom-haberstroh-has-a-smoking) where Tom mentioned that roughly once every few months within a season, the VP of Officiating (currently Monty McCutchen[^1]) will produce a small set of videos that go over POE that relate to certain actions within an NBA game that the officials will pay particular attention to. These points are typically careful considerations about an interpretation of a rule like a pivot foot that relates to traveling or maybe how a shooter follows through on their shot...with their leg.

[^1]: Monty McCutchen, a former NBA referee who was appointed to [the position in December of 2017](https://official.nba.com/monty-mccutchen-vice-president-head-of-referee-development-and-training/). He succeeded Bob Delaney, who stepped down in October of 2017 after 30 years with the NBA. Technically, Monty is "Vice President, Head of Referee Development and Training" but that's a mouthful and VP of Officiating is so much easier to say.

These POE are often [tweeted out](https://x.com/NBAOfficial/status/1707160716162007469) for extra visibility, although it's not really clear how much engagement the [NBA Official](https://x.com/NBAOfficial) account organically creates. But there's always a video accompanied with the POE with a kinda consistent structure to them:

1.  Describe a POE with details on how the referees are instructed to interpret an action/rule.
2.  Display at least one, but potentially multiple, example plays to illustrate an action involving the POE.
    1.  Indicate the player committing the infraction in question as well as the disadvantaged player from the infraction.
    2.  Detail how the action should be ruled.
    3.  State how the action was actually ruled.

There's a lot of information one can grok from these videos. For starters, the amount of time or number of examples shown for a particular POE is easily quantifiable and likely indicates some sort of an *emphasis* on that POE. It is also clear from each of these videos which party is being benefited from the interpretation of a POE, ie is this an example showing that the offense or the defense that will benefit from the interpretation? And a final aspect to glean some information from is that the description of the actual rulings for a POE is similar to the Last Two Minute Reports (L2M) which grade individual actions for NBA games. Every piece of action in an L2M is graded as either a Correct Call (CC), Correct No Call (CNC), Incorrect Call (IC), and Incorrect No Call (INC) which can easily be extended to the example plays described in the Point of Emphasis videos.

We'll dive into these data for the 2023-24 Season in a bit, but before we do that let's get into what I think is a touchy subject within the officiating community: consistency.

# Consistency?

Yes, consistency. You know, when you're emphasizing something it's because you want to hammer home a point about consistently ruling it in a particular way. But how has the NBA officiating done with their POE? Well, let's look backwards a bit.

For the 2022-23 Season, there was a video series involving the [POE with Monty McCutchen](https://www.nba.com/watch/list/collection/points-of-emphasis-2022-23-season) that was pretty similar to what the current season shows. But there exists no POE for the 2021-22 Season but there was a series called [Points of **Education**](https://official.nba.com/video-2021-22-points-of-education/) that Monty McCutchen led. This could still be abbreviated as POE but the E now stands for *education* which might not seem like that big of a change but honestly you're reading this extremely pedantic blog so you already know that I find this to be a substantial change. The series had actually been called Points Of Education dating all the way back to the 2016-17 Season when it was led by Joe Borgia. Joe[^2] retired in [June of 2020](https://pr.nba.com/joe-borgia-retire/) but oversaw a transition in being the figurehead of officiating videos to Monty. Joe also featured in a series that was titled ["Making The Call"](https://official.nba.com/making-the-call/) that began in the 2012-13 NBA Season but stopped being produced around the 2017-18 NBA Season. Monty appears to have taken on a similar series of [Educational Videos](https://videorulebook.nba.com/educational-videos/) that also overlap with POE. But for reference, here is a timeline of the NBA's POE-like videos as far back as I can recall:

[^2]: One aspect of Joe's retirement announcement that I find interesting is: "Borgia is the son of the late Sid Borgia, a legendary referee who officiated in the NBA for 20 years beginning in 1946. Sid Borgia also served as supervisor of officials during the latter years of his career."

| Season  | Series              |     Leader      |                                                                                           website |
|:----------------|:---------------------|:---------------:|----------------:|
| 2023-24 | Points of Emphasis  | Monty McCutchen |                                            <https://official.nba.com/2023-24-points-of-emphasis/> |
| 2022-23 | Points of Emphasis  | Monty McCutchen |                     <https://www.nba.com/watch/list/collection/points-of-emphasis-2022-23-season> |
| 2021-22 | Points of Education | Monty McCutchen |                                     <https://official.nba.com/video-2021-22-points-of-education/> |
| 2020-21 | Points of Education | Monty McCutchen |                                     <https://official.nba.com/video-2020-21-points-of-education/> |
| 2019-20 | Points of Education | Monty McCutchen |                                     <https://official.nba.com/video-2019-20-points-of-education/> |
| 2018-19 | Points of Education | Monty McCutchen |                                     <https://official.nba.com/2018-19-points-of-education-video/> |
| 2017-18 | Points of Education |   Joe Borgia    |                                     <https://official.nba.com/2017-18-points-of-education-video/> |
| 2016-17 | Points of Education |   Joe Borgia    |                                           <https://official.nba.com/2016-17-points-of-education/> |
| 2015-16 | Points of Emphasis  |   Joe Borgia    |                                            <https://official.nba.com/2015-16-points-of-emphasis/> |
| 2014-15 | Points of Emphasis  |   Joe Borgia    |                        <https://official.nba.com/2014-15-nba-officials-points-of-emphasis-video/> |
| 2013-14 | Points of Emphasis  |   Joe Borgia    | <https://official.nba.com/2013-14-nba-officials-points-of-emphasis-and-new-rule-interpretations/> |
| 2012-13 | Rule Authority      |   Joe Borgia    |                                                  <https://official.nba.com/nba-rule-authority-3/> |
| 2011-12 | Points of Emphasis  |   Joe Borgia    |                                        <https://official.nba.com/officiating-points-of-emphasis/> |

Full disclosure: while I have watched all of these video series they were a long time ago and I did not track any of the content within them. I only have vague recollections of what they detailed, so I cannot say anything about a change in interpretation of rules although I highly suspect there have been changes in how certain aspects have been ruled or emphasized.

# POE Analytics - 2023-24

Going back to the structure of POE videos, I have dissected each individual action mentioned in the 2023-24 POE videos. For the 2023-24 Season, there were releases of this series in [September](https://official.nba.com/2023-24-points-of-emphasis/), [December](https://official.nba.com/update-2023-24-points-of-emphasis/), [January](https://official.nba.com/update-2023-24-points-of-emphasis-2/), and [February](https://official.nba.com/update-2023-24-points-of-emphasis-3/) so far. Let's see how the raw mentions bare out:

## Total Emphasis

```{r, echo=FALSE}
poe_data |> 
  group_by(date, emphasis) |> 
  tally() |> 
  mutate(favors = "total") |> 
  replace_na(list(n = 0)) |>
  pivot_wider(names_from = date, values_from = n) |> 
  knitr::kable()
```

In terms of raw mentions, I don't see much of a change in which POE have more mentions throughout the Season. Sure, flopping wasn't talked about much at the beginning and the emphasis on bench conduct has rapidly deteriorated from the first release. But aside from that? It's pretty ho-hum for particular emphasis. At this point, grading the NBA personnel that Ethan talked to I would say they have a point that there was no appreciable change in how the NBA Officials treated particular violations. But maybe the devil is in the details and Monty was signaling changes by the types of plays he was using as an example in the POE videos? We can analyze this by looking at what the points were emphasizing.

## Emphasis by Favorability

```{r, echo=FALSE}
poe_data |> 
  group_by(date, emphasis, favors) |> 
  tally() |> 
  ungroup()|>
  complete(emphasis, favors, date)|>
  replace_na(list(n = 0)) |>
  pivot_wider(names_from = date, values_from = n) |> 
  mutate(favors = factor(favors, levels = c("offense", "defense",
                                            "stars", "nonstars"))) |>
  arrange(emphasis, favors) |> 
  knitr::kable()
```

OK so some of these POE don't really have an offense/defense favorability but instead are related to a star player versus a regular Jalen. This would be the case for issues like respect for the game where examples are of a certain player making a potentially demonstrative gesture towards an official. For example, there was a play where Kevin Durant made a gesture towards an official that warranted a technical foul. In a similar example, it was Marcus Morris Sr. making a gesture towards an official.

But the large takeaway here for me is that there's no appreciable difference in POEs over time towards or away from offense/defense. They seem to be mostly in line with what the POEs started with. So maybe the issue is that Monty is making reference to officials having previously mis-applied rules and he wants to demonstrate that in his examples? I have that information too.

## Emphasis by Grade

```{r, echo=FALSE}
poe_data |> 
  filter(!is.na(grade), grade != "unclear") |>
  group_by(date, emphasis, grade) |> 
  tally() |> 
  ungroup()|>
  complete(emphasis, grade, date)|>
  replace_na(list(n = 0)) |>
  pivot_wider(names_from = date, values_from = n) |> 
  mutate(grade = factor(grade,
                        levels = c("CC", "CNC", "IC", "INC"))) |>
  arrange(emphasis, grade) |> 
  knitr::kable()

```

Every example can be broken down into the CC/CNC/IC/INC categories. And for the most part, they're pretty consistent across time as well. Which largely conforms to what Ethan has been reporting from NBA personnel, nothing in the POEs demonstrate a stark change in **how** games are going to be officiated.

Very interesting topic indeed.

OK now here's the [raw data](poe_2023_24.csv):
